Quantitative Evaluation of Touch-Screen Performance with Transition Matrix Measures

### 2015-09-10 flux-entropy-paper/flux-entropy-paper.tex
ben edits, add some questions/comments to beginning

\begin{abstract}
  HCI systems for supporting collaborative creativity are often
  evaluated with user ratings and interviews. However, such data can
  be impractical to collect in-the-wild, such as during musical
  performances. In this paper we consider measures of gestural
  transitions during collaborative interactions with musical touch
  screen interfaces. The performers' interactions are characterised as
  sequences of discrete gestures and summarised in a transition matrix
  for each performance. Using a corpus of more than 140 performances,
  rehearsals, and demonstrations, we find that two matrix measures,
  flux and entropy, vary significantly with different kinds of
  in-the-wild interactions. Further, using performers' ratings
  collected during a lab-study we find that these measures have a
  significant relationship with the perceived effect of the interface
  on performance. Thus transition analysis, flux and entropy, can be
  used as a proxy for performer ratings during in-the-wild interaction.
\end{abstract}

@@

@@

Finally, we report a quantitative analysis of the transition matrices
from our experimental protocols, and describe the quantities
\emph{flux} and \emph{entropy} which can be used to characterise the
gestural activity of the musicians during a performance. We found a
significant effect of the app on these measures, allowing us to
observe quantitatively the performers' dissatisfaction with a
particular interface. We also investigate the significant relationship
between both of these measures and the survey responses, which
suggests that performers rate interaction with our agent and apps more
highly when engaged in adventurous, exploratory performance.

\section{Related Work}

\begin{itemize}
\item Using Markov chains and transition matrices for usability
  modelling~\cite{Thimbleby:2001kq,Thimbleby:2004fj}
\item coding livecoding~\cite{Swift:2014tya}
\end{itemize}

@@

It has been recognised that collaborative interactions such as
improvised musical performances can be modelled as a sequence of
states. In the following section we present a method for analysing the
transitions between adjacent states in such a sequence and derive two
high-level quantities that we use to compare musical performances
under different circumstances. Suppose that an collaborative
interaction can be characterised by a set of interaction states $S$
with $m$ members. Each participant's activity can then be represented
as a sequence:
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
Where each $X_i$ is a member of $G$. The transitions between different
interaction states can be summarised in a transition matrix as if the
sequence $X_n$ were a first-order Markov chain. Let $N_{ij}$ be the
number of times that state $j$ follows state $i$. Each participant's
transition matrix, $P$, will then be an $m \times m$ matrix with
entries calculated thus
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the transition matrix for each performer.
The matrix is normalised so that the sum of all elements will be one
(rather than the sum of each row being one, as for a right-stochastic matrix).

To compare the transition matrices of different interactions we use
two high-level quantities, ``flux'' and ``entropy''. The flux of a
sequence is defined as the ratio of changed transitions (e.g.
$A \rightarrow B$), to self transitions (e.g. $A \rightarrow A $). The
the flux of a transition matrix $P$ can easily be calculated:
\begin{equation}
  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
\end{equation}
Where the trace of P, $tr(P)$ is given by the sum of the diagonal
entries of $P$.
% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
% $P$.
Intuitively, flux is a measure of how frequently a group changes state. Flux
returns a value in the interval $[0,1]$ and will return 0 when
participants never change state, and 1 when participants never stay on
the same state for two measurements in a row.

@@

\section{A Corpus of Touch-Screen Interactions}

@@

In the following sections, we will use transition matrices and the
flux measure to analyse a corpus of data collected as part of our
research into collaborative interactions with musical touch-screen
tablet interfaces.

Over a period of more than two years (April 2013 - June 2015), we have
collected data from $150$ sessions where various participants used
musical touch-screen interfaces on Apple iPad devices. These
interfaces all shared a common mode of interaction where the majority
of the screen could be used for free form touch interactions; tapping
produced short sounds, while swiping or swirling produced long sounds
with volume proportional to the velocity of the moving touch point.
All touch interactions from each participant in a session was recorded
by sending data to a central server application. This data was stored
in CSV files recording the date and time, X, and Y locations,
velocity, and a unique device identifier for each touch event detected
by the iOS operating system.

@@

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figures
\ref{fig:flux-entropy-boxplot} and
\ref{fig:flux-entropy-distribution}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values. In the following
analysis, only the 95 performative sessions were considered (i.e. not
the demonstration or data collection contexts).

@@

\begin{itemize}
\item todo: need to add overall contrasts column to do pairwise t-tests.
\end{itemize}

@@

It is clear that flux and entropy can detect the variation in
performance styles in between performances of different types, but can
they be used to understand some of the internal structure of
performances? A simple model for understanding temporal art-forms is
the ``beginning, middle, end'' structure. To investigate how the
structure fits session in our corpus, we can divide the
gestural-sequences of each session into thirds and calculate the
transition matrices given by each section.

Figures \ref{fig:type-section-flux-entropy} shows that the flux and
entropy of compositions are much more stable with respect to section
than for improvisations. The stability of compositions is to be
expected since performers are expected to use the same gestures in
each rehearsal and performance of a composition, it may be that
different kinds of compositions could result in different structural
differences. Both measures have a downward trend through sections in
improvisations with Kruskal-Wallis tests showing a significant effect
of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
by performers entering an exploration stage at the start of an
improvisation, where they change gesture frequently to experiment with
different ideas. At the end of an improvisation, performers may change
gesture less frequently while ``winding up'' the performance.

@@

While the interaction session in our corpus have many different
contexts, they are all part of a program of artistic and HCI research
that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
shows the values of flux and entropy for each session in the corpus.
Over the course of this research program, different ensembles and
different activities have been emphasised, some of these differences
can be seen as clusters of performances with similar flux and entropy
in the two plots. For instance, the earliest series of rehearsals and
performances had flux between 0.05 and 0.15, while the period of
activity after July 2014 had flux between 0.1 and 0.25. This change
corresponded with an intensive series of study sessions and rehearsals
where the participants developed a more fluctuating style of
performance. The entropy between these two periods, however, remains
at a similiar level, showing the the space of available gestures that
the performers explored was more constant.

@@

In this paper we have extended the transition matrix methodology
previously presented at CHI with two simple matrix measures, flux and
entropy, both of which have been demonstrated to have significant
discriminatory power with regards to a corpus of 150 collaborative
touch-screen interactions. Not only do these measures differentiate
between different types of sessions and the internal structure of
these sessions, but they have explanatory power that has thrown new
light on the musical interactions under examination. HCI researchers
are increasingly concerned with interactions that happen in-the-wild,
where the interactions of many users may be collected without the
opportunity to survey or interview users. We suggest that analysing
interaction-state transitions using matrices and simple measures can
lead to new insights into HCI interactions outside the lab.

>>>>

\begin{abstract}
  HCI systems for supporting collaborative creativity are often
  evaluated with user ratings and interviews. However, such data can
  be impractical to collect in-the-wild, for example during musical
  performances. In this paper we show how a transition matrix based
  post-hoc analysis of logged interaction from a long program of
  experiments in group musical creativity can shed light on the nature
  of the group interaction in-the-wild. Using a corpus of more than
  140 performances, rehearsals, and demonstrations over more than two
  years, we find that two matrix measures, flux and entropy, can be
  used as a proxy for performer ratings during in-the-wild
  interaction.
\end{abstract}

@@

\section{General comments}

\begin{itemize}
\item we should be clear (and careful) where we're talking about
  groups TMs, and where we're talking about individual performers
\item what's the terminology we're using for a ``performance''? Is it
  ``performance'', or is that too music-specific? ``Session''?
  Alternatives are ``activity'', or ``interaction'', but they're so
  general\ldots This decision has implications for our other
  terminology: participant vs performer vs musician, etc.
\item Title: maybe ``Traces through time: making sense of longitudinal
  open-ended interaction data'' or some such? I think that's our big
  contribution: it takes the in-the-wild, practitioner-driven research
  program stuff and tries to make some quantitative sense out of it
  post-hoc.
\end{itemize}

@@

Ideas: the in-the-wild nature, and the different contexts - both the
fact that they are different, and the fact that their differences make
collecting certain data/responses easier or harder

@@

Open-ended collaborative interactions such as improvised musical
performances can be modelled as a sequence of
% could also use Thimbleby:2004fj
states\cite{Thimbleby:2001kq,Swift:2014tya}. In this section we use
transition matrix (TM) statistics to analyse the transitions between
adjacent states in such a sequence. We then present two high-level
derived quantities---\emph{flux} and \emph{entropy}---which can be
used to compare musical performances under different circumstances.

Suppose that each participant's activity in a collaborative
interaction can at any given time be classified as belonging to one of
$m$ distinct \emph{states}. Over the course of the activity, each
participant $X$ generates a length $N$ sequence of these interaction
states
\begin{equation}
 X_n \hskip 2em n = 1, \ldots, N
\end{equation}
The transitions between these interaction states can be summarised in
a transition matrix (TM) as if the sequence $X_n$ were a first-order Markov
chain. Let $N_{ij}$ be the number of times that state $j$ follows
state $i$. Each participant's TM $P$, will then be an
$m \times m$ matrix with entries
\begin{equation}
  p_{ij} = \frac{N_{ij}}{\sum_i \sum_j N_{ij}}
\end{equation}
Each entry of $P$ will be such that the transition probability from
state $i$ to state $j$ is given by the entry in the $i^{th}$ row and
the $j^{th}$ column. The transition activity of the whole ensemble can
be calculated by averaging the TM for each performer.
The matrix can be normalised so that the sum of all elements is equal
to one.
% (rather than the sum of each row being one, as for a right-stochastic matrix)

If there is an (automated) classification process for determining each
participant's state sequence during an activity, then it is a simple
process to generate a TM for each participant and for the group as a
whole. The challenge is to understand what the TM reveals about the
group interaction, and to allow for comparisons between sessions and
groups.

To compare the TMs of different interactions we use two high-level
quantities, ``flux'' and ``entropy''. The flux of a sequence is
defined as the ratio of change transitions (e.g. $A \rightarrow B$
where $A \neq B$), to self transitions (e.g. $A \rightarrow A $). The
flux of a TM $P$ is given by
\begin{equation}
  \mathrm{flux}(P) = 1 - \mathrm{tr}(P)
\end{equation}
where the trace $\mathrm{tr}(P)$ is the sum of the diagonal entries of
$P$.
% Where $\|P\|_1 = \sum_{i,j}|p_{ij}|$ is the element-wise 1-norm of $P$
% and $\mathrm{diag}(P)$ is the vector of the main diagonal entries of
% $P$.
Intuitively, flux is a measure of how frequently the participant/group
changes state. Flux returns a value in the interval $[0,1]$ and will
return 0 when participants never change state, and 1 when participants
never stay in the same state for two measurements in a row.

@@

\section{Making sense of two year's worth of touch-screen
  interactions}

@@

% TODO maybe a ref to the CHI paper here?

% TODO I think this ``scene-setting'' stuff could actually go in the
% first section of the paper, with the TM stuff following.  But I'm
% open to push-back on that.

Over a period of more than two years (April 2013 - June 2015), we have
collected data from $150$ group musicmaking sessions where
participants used musical touch-screen interfaces on Apple iPad
devices. These interfaces all shared a common mode of interaction
where the majority of the screen could be used for free form touch
interactions; tapping produced short sounds, while swiping or swirling
produced long sounds with volume proportional to the velocity of the
moving touch point. Through all of these performances every touch
interaction---every touch down, dragged and up---was timestamped and
logged to disk for later analysis

% TODO say something about how it was both your research project and
% your creative practice. That's why it encompassed all these
% different contexts.

@@

% TODO again, I think this bit would be beefed-up if we included the
% fact that it's already been published.  And it also means we don't
% have to go into as much detail about how we did it - the main thing
% is that we've got a sensible (and peer-reviewed!) way of taking the
% touch interaction data and generating gesture sequences (and
% therefore TMs)

@@

The distribution of flux and entropy with respect to session contexts,
types and instrumentation are shown in Figures
\ref{fig:flux-entropy-boxplot} and
\ref{fig:flux-entropy-distribution}. A three-way ANOVA procedure was
performed on the data set to find whether the metatdata could
significantly predict the entropy and flux values.
% TODO we haven't explained what ``the 95 performative sessions'' are,
% have we?
In the following analysis, only the 95 performative sessions were
considered (i.e.~not the demonstration or data collection contexts).

@@

\begin{itemize}
\item TODO: need to add overall contrasts column to do pairwise t-tests.
\end{itemize}

@@

It is clear that flux and entropy can detect the variation in
performance styles in between performances of different types, but can
they be used to understand some of the internal structure of
performances? A simple model for understanding temporal art-forms is
the ``beginning, middle, end'' structure. To investigate how this
structure fits our corpus, we can divide the gestural-sequences of
each session into thirds and calculate the transition matrices given
by each section.

Figure~\ref{fig:type-section-flux-entropy} shows that the flux and
entropy of compositions are much more stable with respect to section
than for improvisations. The stability of compositions is to be
expected since performers are expected to use the same gestures in
each rehearsal and performance of a composition, it may be that
different kinds of compositions could result in different structural
differences. Both measures have a downward trend through sections in
improvisations with Kruskal-Wallis tests showing a significant effect
of section on flux ($\chi^2(2)=5.8, p=0.05$). This could be explained
by performers entering an exploration stage at the start of an
improvisation, where they change gesture frequently to experiment with
different ideas. At the end of an improvisation, performers may change
gesture less frequently while ``winding up'' the performance.

@@

While the interaction session in our corpus have many different
contexts, they are all part of a program of artistic and HCI research
that has evolved over time. Figure \ref{fig:flux-entropy-through-time}
shows the values of flux and entropy for each session in the corpus.
Over the course of this research program, different ensembles and
different activities have been emphasised, some of these differences
can be seen as clusters of performances with similar flux and entropy
in the two plots. For instance, the earliest series of rehearsals and
performances had flux between 0.05 and 0.15, while the period of
activity after July 2014 had flux between 0.1 and 0.25. This change
corresponded with an intensive series of study sessions and rehearsals
where the participants developed a more ``fluctuating'' style of
performance. The entropy between these two periods, however, remains
at a similar level, showing the the space of available gestures that
the performers explored was relatively stable.

@@

In this paper we have extended our previous work on transition
matrices with two simple matrix measures, flux and entropy, both of
which have been demonstrated to have significant discriminatory power
with regards to a corpus of 150 collaborative touch-screen
interactions. Not only do these measures differentiate between
different types of sessions and the internal structure of these
sessions, but they have explanatory power that has thrown new light on
the musical interactions under examination. HCI researchers are
increasingly concerned with interactions that happen in-the-wild,
where the interactions of many users may be collected without the
opportunity to survey or interview users. We suggest that analysing
interaction-state transitions using matrices and simple measures can
lead to new insights into HCI interactions outside the lab.
